---
title: "24h EPG Data Analysis and Figures"
author: "Daniel Kunk"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
---
# 1. General Data Import and analysis

Loading our 24 h data with two treatments of LD and LL
```{r}
# load the data
LD_LL_24 = read.csv(
  "./data/24h_ShinyAppParameterDataExport_2026-04-07 19_48_18.csv")
```

Making a set of key waveforms that will be analyzed. See discoEPG shinyapp/r-package for full variable list.
We can also add a set of values to each name in the vector to use for unit conversions later, rather than on the fly.

```{r}
# load a vector of target variables for future use. 
target_variables = c(
  "Total duration of np" = "h",
  "Total duration of pd" = "min",
  "Total duration of C"  = "h",
  "Total duration of E1" = "h",
  "Total duration of E2" = "h",
  "Total duration of F"  = "min",
  "Duration of G"        = "h",
  "Number of np"         = "count",
  "Number of pd"         = "count",
  "Number of C"          = "count",
  "Number of E1"         = "count",
  "Number of E2"         = "count",
  "Number of F"          = "count",
  "Number of G"          = "count"
)
```

We can take a peek at the distribution of our data.

```{r}
library(dplyr)
library(ggplot2)
library(ggridges)

# univariate histograms for all target variables
# first grabbing a small target dataset
target_data = LD_LL_24 %>%
  filter(description %in% names(target_variables)) %>%
  mutate(value = as.numeric(value))

# plotting all histograms at once using facet_wrap
ggplot(target_data, aes(x = value, fill = treatment)) +
  geom_histogram(bins = 30, color = "black") +
  theme_classic() +
  facet_wrap(~ description, scales = "free") + 
  labs(
    x = "Value",
    y = "Frequency"
  ) +
  theme(legend.position = "none")

# plotting all density plots at once using facet_wrap
ggplot(target_data, aes(x = value,, y = treatment, fill = treatment)) +
  geom_density_ridges(alpha = 0.5, color = "black", scale = 1.5) +
  theme_classic() +
  facet_wrap(~ description, scales = "free") + 
  labs(
    x = "Value",
    y = "Frequency"
  ) +
  theme(legend.position = "none")
```

We can see that we have less overall samples than the 8h data and that we see the typical, wonky distributions. Not normal, lots of variation from aphid to aphid.

Likely, we want to use gamma GLMs with appropriate zero-inflation/over-dispersion terms as necessary for our duration behaviors. In addition to poisson and negative-binomial models for the count data.

We can see what the summary values of our data are. 

```{r}
# first lets see if any of our duration and count variables have zeros.
summary_24h = LD_LL_24 %>%
  filter(description %in% names(target_variables)) %>%
  group_by(description, treatment) %>%
  summarise(
    max = max(value),
    min = min(value),
    n = n(),
    mean = mean(value),
    sd = sd(value),
    median = median(value),
    mode = DescTools::Mode(value)[1],
    .groups = "drop_last"
  )

# save to csv for supplementary tables
### write.csv(summary_24h, file = "./suppData/24h_EPG-SummaryStatistics")

```

We can see that there are lots of zeros for the count data. We can also see the mode for the count data.
For the duration data we can see that the only zeros are found in G and F which makes sense. We can also see that the modes for the duration data are all NA, this is because all these numbers are unique and there isn't any most frequent number.

With this knowledge we know that a gamma distribution will work for any of the duration variables except for F and G, in these cases we could use a hurdle model or better yet a Tweedie model. For the count data, we can likely get by with a negative binomial distribution for all of our variables.


It's always a good idea to start out with the total duration and number of C. Let's build some simple models and check the fit + assumptions.
```{r}
# a small modelling sandbox

# load required modeling and diagnostic packages
library(dplyr)
# for modelling
library(MASS)     
# diagnostics
library(DHARMa)    

# prep the data 
sandbox_data = LD_LL_24 %>%
  filter(description %in% names(target_variables)) %>%
  mutate(value = as.numeric(value))

# 1. gamma model (durations > 0)
# "Total duration of C"

data_gamma = sandbox_data %>% filter(description == "Total duration of C")

# fit gamma with log link
m_gamma = glm(value ~ treatment, family = Gamma(link = "log"),
              data = data_gamma)

# evaluate gamma Model
print(summary(m_gamma), show.residuals = TRUE)

# base diagnostics
par(mfrow = c(2,2))
plot(m_gamma, main = "gamma: Duration of C")

# DHARMa 
res_gamma = simulateResiduals(fittedModel = m_gamma, plot = TRUE)

# 2. tweedie (durations with exact 0s)
# "Total duration of F"

data_tweedie = sandbox_data %>% filter(description == "Total duration of F")

# fit tweedie model
# var.power = 1.5 is the standard compound poisson-gamma dist. (handles 0s)
# link.power = 0 acts as a log link function
m_tweedie = glm(value ~ treatment, 
                family = tweedie(var.power = 1.5, link.power = 0), 
                data = data_tweedie)

# eval tweedie Model

print(summary(m_tweedie), show.residuals = TRUE)

# base diagnostics
par(mfrow = c(2,2))
plot(m_tweedie, main = "tweedie: Duration of F")

# DHARMa
res_tweedie = simulateResiduals(fittedModel = m_tweedie, plot = TRUE)

# 3. neg binom (for 0 counts)
# "Number of C"

data_negbin = sandbox_data %>% filter(description == "Number of C")

# fit negative binomial
m_negbin = glm.nb(value ~ treatment, data = data_negbin)

# eval nb model
print(summary(m_negbin), show.residuals = TRUE)

# base diagnostics
par(mfrow = c(2,2))
plot(m_negbin, main = "NegBin: Number of C")

# DHARMa
res_negbin = simulateResiduals(fittedModel = m_negbin, plot = TRUE)

# reset plot window to default
par(mfrow = c(1,1))
```

We can see that the gamma distribution will be just fine for everything except F and G where the Tweedie distribution GLM is useful.
Further we can see that we have no problem using the negative binomial for the counts. Some could likely use the poisson due to not as much overdispersion as with our other dataset, but the fit will be generally the same.

# 2. Making final box-plots.

Here we can make some finalized figures inside a small list. Here we are also using the target variables variable from above.

```{r}
# load packages if needed
library(ggplot2)
library(dplyr)

# grab the data as above. Using names for named vectors.
target_data = LD_LL_24 %>%
  filter(description %in% names(target_variables)) %>%
  mutate(value = as.numeric(value))

# create an empty list to add plots to
boxplot_list = list()

# loop through the the target_variables names specified above
for (var in names(target_variables)) {
  
  # extract the target unit for this specific variable
  unit = target_variables[[var]]
  
  # subset data for the current behavior/waveform
  plot_data = target_data %>% 
    filter(description == var)

  # setup conditional logic for unit conversions and y-axis labeling
  if (unit == "h") {
    plot_data$value = plot_data$value / 3600
    y_label = paste(var, "(hours)")
  } else if (unit == "min") {
    plot_data$value = plot_data$value / 60
    y_label = paste(var, "(minutes)")
  } else if (unit == "sec") {
    # no division for seconds
    y_label = paste(var, "(seconds)")
  } else {
    # default for "count" or anything else
    y_label = paste(var, "(count)")
  }
  
  # generate the plots
  p = ggplot(plot_data, aes(x = treatment, y = value, fill = treatment)) +
    geom_violin() + 
    geom_jitter(width = 0.05) + 
    geom_boxplot(width=0.1, color="black", alpha=0.2, outlier.shape = NA) +
    theme_classic() +
    labs(
      x = NULL,
      y = y_label # injects the dynamic label here based on conditional above
    ) +
    theme(
      legend.position = "none"
    )
  
  # store the plot in our list using the variable name as the key
  boxplot_list[[var]] = p
  
  # print plots to viewer
  print(p)
}
```

Those plots are nice above but below we will create plots that take into account the p-values from our statistical analysis.

```{r}
# if not already loaded
library(ggplot2)
library(dplyr)
# for modelling
library(MASS)
# make prettier final plots
library(ggprism)

# clean the data
target_data = LD_LL_24 %>%
  filter(description %in% names(target_variables)) %>%
  mutate(value = as.numeric(value))

# create an empty list to store our final plots
boxplot_list = list()

# start the for loop to generate the plots.

# loop through the variables
for (var in names(target_variables)) {
  
  unit = target_variables[[var]]
  plot_data = target_data %>% filter(description == var)
  
  
  # unit conv. and axis labels
  if (unit == "h") {
    plot_data$value = plot_data$value / 3600
    y_label = paste(var, "(hr)")
  } else if (unit == "min") {
    plot_data$value = plot_data$value / 60
    y_label = paste(var, "(min)")
  } else if (unit == "sec") {
    y_label = paste(var, "(sec)")
  } else {
    y_label = paste(var)
  }
  
  # starting modelling and p-value extraction
  p_val = NA
  sig_label = "ns"
  
  tryCatch({
    # 1. neg binom for all count data
    if (unit == "count") {
      mod = glm.nb(value ~ treatment, data = plot_data)
      
    # 2. tweedie for specific zero-inflated durations F and G
    } else if (var %in% c("Total duration of F", "Duration of G")) {
      mod = glm(value ~ treatment, family = tweedie(var.power = 1.5, 
                link.power = 0), data = plot_data)
      
    # 3. gamma for all other duration vars
    } else {
      mod = glm(value ~ treatment, family = Gamma(link = "log"), 
                data = plot_data)
    }
    
    # extract the summary table
    coefs = summary(mod)$coefficients
    
    # dynamically find the row name for the coefficient "treatmentLL"
    term_name = grep("treatment", rownames(coefs), value = TRUE)[1]
    
    # the 4th column in a GLM summary matrix is the p-value
    if (!is.na(term_name)) {
      p_val = coefs[term_name, 4] 
    }
    
    # convert p-value to asterisks
    sig_label = case_when(
      is.na(p_val)  ~ "Error",
      p_val < 0.001 ~ "***",
      p_val < 0.01  ~ "**",
      p_val < 0.05  ~ "*",
      TRUE          ~ "ns"
    )
    
  }, error = function(e) {
    # if a model fails to converge, print an error on the plot; debug
    sig_label <<- "Model Error" 
  })
  
  # starting the plot p-value annotation settings
  # find the maximum value to guide bracket placement
  max_y = max(plot_data$value, na.rm = TRUE)
  
  # set heights based on a percentage of the max value 
  bracket_y = max_y + (max_y * 0.08) 
  text_y = bracket_y + (max_y * 0.04) 
  upper_limit = text_y + (max_y * 0.05) # keep the text going oob
  
  
  # make the plot
  p = ggplot(plot_data, aes(x = treatment, y = value, fill = treatment, 
                            shape = treatment)) +
    geom_violin(color = NA) + 
    geom_boxplot(width = 0.5, color = "grey10", outlier.shape = NA, 
                 alpha = 0.5) +
    geom_jitter(width = 0.1, size = 3, height = 0) +
    scale_shape_manual(
      values = c("LD" = 21, "LL" = 22)) +
    scale_fill_manual(values = c("LD" = "#0077BB", "LL" = "#EE7733")) +
    # now adding the significance line and text
    annotate("segment", x = 1, xend = 2, y = bracket_y, 
             yend = bracket_y, size = 0.5) +
    annotate("text", x = 1.5, y = text_y, label = sig_label, 
             size = 6, fontface = "bold") +
    theme_prism() +
    labs(x = NULL, y = y_label) +
    coord_cartesian(ylim = c(0, upper_limit)) + # prevents text cutoff
    theme(
      legend.position = "none"
    )
  
  # store the plot to list and then print
  boxplot_list[[var]] = p
  print(p)
}
```

Now that all of these boxplots are made, we want to be able to grab them using ggsave to make high quality svgs or pdfs.

```{r}
# save a specific plot as a PDF, use the list
      # ggsave(
      #   filename = "./figures_24h/Total_duration_of_C.pdf",
      #   plot = boxplot_list[["Total duration of C"]],
      #   width = 5,   # adjust h and w accordingly...
      #   height = 5,
      #   device = "pdf"
      # )
```

Looping to make svgs.

```{r}
# global width and height
plot_width = 5
plot_height = 5

# loop through the boxplot list and save each plot to pdf
for (var_name in names(boxplot_list)) {
  
  # clean up the name for the file system (spaces to underscores)
  safe_name = gsub(" ", "_", var_name)
  
  # construct the full file path
  file_path = paste0("./figures_24h/boxplots/", safe_name, ".svg")
  
  # save
        # ggsave(
        #   filename = file_path,
        #   plot = boxplot_list[[var_name]],
        #   width = plot_width,
        #   height = plot_height,
        #   device = "svg"
        # )
  # debug print to see all file saved...
  print(paste(safe_name, "saved to", file_path))
}
```

Making multi-plots for pub. with patchwork.

```{r}
library(patchwork)

# target variables
specific_vars = c("Total duration of C", "Total duration of E1", 
                  "Total duration of E2", "Number of E1")

# make sure they are available in the list
available_vars = intersect(specific_vars, names(boxplot_list))

if (length(available_vars) > 0) {
  
  # extract the targeted plots
  multi_plots_24h = boxplot_list[available_vars]
  
  # use wrap_plots to form a 2-column grid
  combined_grid_24h = wrap_plots(multi_plots_24h, ncol = 2) &
    plot_layout(axis_titles = "collect_x") & 
    plot_annotation(tag_levels = 'A')
  
  
  # save as svg
  # ggsave(
  #   filename = "./figures_24h/multiplots/variables_durations_and_counts_24h.svg",
  #   plot = combined_grid_24h,
  #   width = 9,      # Dimensions fit nicely for a square-ish 2x2 layout
  #   height = 8,
  #   device = "svg"
  # )
  # show plot
  print(combined_grid_24h)
  # echo done
  print("Saved 24h Specific Multiplot successfully!")
}
```

# 3. Making waveform signal and time series frequency plots.

First we can start out with time series frequency plots. These will summarize the most frequent waveform at each time step. This is objective data and nothing is being modeled here like was done in the past.

```{r}
# for loading ana files, assigning treatment, and plotting.
library(discoEPG)
# load if not done already (dplyr, ggplot2, etc.)
library(tidyverse)
library(ggprism)

# define path to ana files
ana_dir_24 = "../EPG_data/24h_data/24h_ana/"
# grab file names
filenames = list.files(path = ana_dir_24, pattern = ".ANA")

# save a small named vector of waveform labels
waveform_labels = c("np" = 1, "C" = 2, "E1e" = 3, "E1" = 4, 
                    "E2" = 5, "F" = 6, "G" = 7, "pd" = 8, "end" = 12)

# nolad 24h data w/ discoEPG
anaDatList_24 = discoEPG::combine_ana(directory = ana_dir_24,
                                      filenames = filenames,
                                      validate = TRUE,
                                      waveform_labels = waveform_labels)
# extract the data
anaData_24 = anaDatList_24$data

# we will assign treatments to the data based on ch number
# make a list of file names to assign treatments to
allfiles = unique(anaData_24$filename)
LD_treat_files = grep("ch[1-4]", allfiles, value = TRUE)
LL_treat_files = grep("ch[5-8]", allfiles, value = TRUE)

# assign treatments
anaData_24$treatment = NA # debug
# LD
anaData_24$treatment[anaData_24$filename %in% LD_treat_files] = "LD"
# LL
anaData_24$treatment[anaData_24$filename %in% LL_treat_files] = "LL"

# check for missing assignments, debug
if(any(is.na(anaData_24$treatment))) {
  warning("Some files were not assigned a treatment group! Check your filename patterns.")
}

# grab the data output from ana_to_timeseries_proportion()
# use it to make the summary plots faceted by LD and LL
combined_prop_24 = anaData_24 %>%
  group_split(treatment) %>%
  map_dfr(function(df) {
    tr = unique(df$treatment)
    
    # extract data from the discoEPG output list
    res = discoEPG::ana_to_timeseries_proportion(df, waveform_labels)
    
    # return just the data, add treatment column back in
    res$data %>%
      mutate(treatment = factor(tr, levels = c("LD", "LL")))
  })

# set plot colors (colorblind safe - paul tol bright palette) 
behavior_colors = c(
  "NP" = "black", "C"  = "#0077BB", "E1" = "#33BBEE",
  "E2" = "#009988", "F"  = "#EE7733", "G"  = "#CC3311", "PD" = "#EE3377"
)

# set up a shading layer mapped only to LD facet
shade_df_24 = data.frame(
  treatment = factor("LD", levels = c("LD", "LL")),
  xmin = 16, xmax = 24,
  ymin = -Inf, ymax = Inf
)

# build the facet plot with facet_grid()
faceted_summary_24h = ggplot(combined_prop_24, 
                             aes(x = hours, y = proportion, color = activity)) +
  
  # add shading layer first
  geom_rect(
    data = shade_df_24,
    aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = ymax),
    inherit.aes = FALSE,
    fill = "grey30", alpha = 0.15
  ) +
  
  # smoothing to make pretty...
  stat_smooth(method = "gam", 
              formula = y ~ s(x, bs = "cs", k = 40), 
              se = FALSE, linewidth = 1, alpha = 0.2) +
  
  # grid layout.
  facet_grid(treatment ~ ., scales = "fixed") + 
  
  # theme
  theme_prism() +
  # pretty breaks
  coord_cartesian(xlim = c(0, 24)) +
  scale_y_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1), limits = c(0, 1)) +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) + 
  # add the color palette above set order in legend
  scale_color_manual(
    values = scales::alpha(behavior_colors, 0.6),
    breaks = c('NP', 'C', 'PD', 'E1', 'E2', 'F', 'G')
  ) +
  # add labels
  labs(x = "Zeitgeber Time", y = "Proportion of Aphids", color = "Behavior") +
  # legend on the right, add some space between the two panels
  theme(
    legend.position = "right",
    panel.spacing.y = unit(1.5, "lines")
  ) +
  # make legend line width larger
  guides(color = guide_legend(override.aes = list(linewidth = 2)))

# render the plot in the window, may take a minute 
#print(faceted_summary_24h)


# save the plot as svg. may take a while.
    # ggsave(
    #   filename = "./figures_24h/summary/TimeSeries_Summary_24h.svg",
    #   plot = faceted_summary_24h,
    #   width = 7.5,
    #   height = 6,
    #   device = "svg"
    # )
```

Now we can make the signal plots showing representative waveforms across our two treatments.

```{r}
# first we can make some plots that will summarize EPG signals across treatments
# we can pick a bunch of random recordings across each light treat
# then we can facet them with patchwork to make them look good.
# to do this we can use EPGplotR and a loop to speed things up.

# load packages
library(EPGplotR)
# if not loaded already
library(patchwork)
library(ggplot2)
library(ggprism)

# select the representative files for each treat
rep_files = c(
  # LD Representatives
  "24h_LD_LL_Rep3-ch4", 
  "LD-LL_16_8_0zt-ch3", 
  "LD-LL_16_8_0zt_rep2-ch4", 
  # LL Representatives
  "24h_LD_LL_Rep3-ch5", 
  "LD-LL_16_8_0zt-ch7", 
  "LD-LL_16_8_0zt-ch8"
)

# define directories for data and ana
dxx_dir = "../EPG_data/24h_data/24h/" 
ana_dir = "../EPG_data/24h_data/24h_ana/"

# def new colors.
my_new_colors = c(
  "np" = "grey30",
  "C"  = "#0077BB",
  "E1" = "#33BBEE",
  "E2" = "#009988",
  "F"  = "#EE7733",
  "G"  = "#CC3311",
  "pd" = "#EE3377"
)

# create the empty list to hold the plots
signal_plot_list = list()

# loop through, generate plots, and store them to list
for (file_base in rep_files) {
  
  d01_path = paste0(dxx_dir, file_base, ".D01")
  ana_path = paste0(ana_dir, file_base, ".ANA")
  # small debug to make sure files exist
  if(file.exists(d01_path) & file.exists(ana_path)) {
    
    # load data and annotate with EPGplotR functions
    rec_dt = load_d0x_files(d01_path)
    ana_dt = load_ana_file(ana_path)
    assign_behavior_codes(rec_dt, ana_dt)
    
    # generate the static plot
    p = gg_plot_signal(
      rec_dt, 
      downsample_rate = 10, 
      alpha = 0.8, 
      plot_title = paste("Rec. Name = ", file_base)
                      ) + 
      theme_prism() +
      labs(x = "Zeitgeber Time") +
      # new color. wont see changes on initial print
      # as gg_plot_signal() has a print statement within the function...
      scale_color_manual(values = my_new_colors, name = "Behavior",
                         breaks = c("np", "C", "pd", "E1", "E2", "F", "G"),
                         labels = c("NP", "C", "PD", "E1", "E2", "F", "G")) +
      # rescale and rename x axis
      scale_x_continuous(
        name = "Zeitgeber Time",       
        breaks = seq(0, 24, by = 4),   
        limits = c(0, 24),             
        expand = c(0, 0)               
      ) +
      # adjust plot title size
      theme(plot.title = element_text(size = 8, face = "plain" ),
            legend.position = "none") +
      guides(color = guide_legend(override.aes = list(linewidth = 2)))
    
    # add to the list
    signal_plot_list[[file_base]] = p
    # small debug print statement
    print(paste("Generated plot for:", file_base))
    
  } else {
    warning(paste("Missing D01 or ANA file for:", file_base))
    # above rest of small debug to catch missing files...
  }
}

# stitch into two columns (LD on left, LL on right)
# byrow = FALSE forces patchwork to fill down column 1 before starting column 2
combined_signal_plot = wrap_plots(signal_plot_list, ncol = 2, byrow = FALSE, guides = "collect", axes = "collect") +
  plot_annotation(
    title = 'Representative 24 h EPG Signals',
    subtitle = 'Left Column: LD Treatment | Right Column: LL Treatment',
    theme = theme(
      plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 14, hjust = 0.5)
    )
  )
# quick look to see how it is, will take a minute
#print(combined_signal_plot)

# save, will take a minute
# about the width of US letter with 1 in margins
    # ggsave(
    #   filename = "./figures_24h/signals/Combined_24h_Signals_Faceted.svg",
    #   plot = combined_signal_plot,
    #   width = 7.5,
    #   height = 10,
    #   device = "svg")
```

