---
title: "8h EPG Data Analysis and Figures"
author: "Daniel Kunk"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
---

# 1. Load the data

Here we can load the two datasets, format them, and perform some summary statistics.

```{r}
library(dplyr)
# load the data and add a light treatment column

LD_8h = read.csv(
  "./data/LD_ShinyAppParameterDataExport_2026-04-07 20_03_26.csv") %>% 
  mutate(light = "LD")

LL_8h = read.csv(
  "./data/LL_ShinyAppParameterDataExport_2026-04-07 20_06_57.csv") %>% 
  mutate(light = "LL")

# join the two dataframes into one
LD_LL_8h = rbind(LD_8h, LL_8h)

# clean memory
rm(LD_8h, LL_8h)

# factorize treat
LD_LL_8h$treatment = factor(LD_LL_8h$treatment, 
                            levels = c("ZT0", "ZT8", "ZT16"), 
                            ordered = F)

# and light
LD_LL_8h$light = factor(LD_LL_8h$light, 
                        levels = c("LD","LL"), 
                        ordered = F)
```

We can make a named vector of the variables that we care about.

```{r}
# load a vector of target variables for future use. 
target_variables = c(
  "Total duration of np" = "h",
  "Total duration of pd" = "min",
  "Total duration of C"  = "h",
  "Total duration of E1" = "h",
  "Total duration of E2" = "h",
  "Total duration of F"  = "min",
  "Duration of G"        = "h",
  "Number of np"         = "count",
  "Number of pd"         = "count",
  "Number of C"          = "count",
  "Number of E1"         = "count",
  "Number of E2"         = "count",
  "Number of F"          = "count",
  "Number of G"          = "count"
)
```

Some slight EDA.

```{r}
library(ggplot2)
library(ggridges)
# univariate histograms for all target variables
# first grabbing a small target dataset
target_data = LD_LL_8h %>%
  filter(description %in% names(target_variables)) %>%
  mutate(value = as.numeric(value))

# plotting all histograms at once using facet_wrap
ggplot(target_data, aes(x = value, fill = treatment)) +
  geom_histogram(bins = 30, color = "black", alpha = 0.5) +
  theme_classic() +
  facet_wrap(vars(description, light), scales = "free") + 
  labs(
    x = "Value",
    y = "Frequency"
  ) +
  theme(legend.position = "none")

# using a ridges geom as it looks a bit better than overlapped histograms.
ggplot(target_data, aes(x = value, y = treatment, fill = treatment)) +
  geom_density_ridges(alpha = 0.5, color = "black", scale = 1.5) +
  theme_classic() +
  facet_wrap(vars(description, light), scales = "free") + 
  labs(
    x = "Value",
    y = "Treatment" 
  ) +
  theme(legend.position = "none")
```

Some more EDA to get summary statistics on the data

```{r}
# using both light and treatment
summary_8h = LD_LL_8h %>%
  filter(description %in% names(target_variables)) %>%
  group_by(description, treatment, light) %>%
  summarise(
    max = max(value),
    min = min(value),
    n = n(),
    mean = mean(value),
    sd = sd(value),
    median = median(value),
    mode = DescTools::Mode(value)[1],
    .groups = "drop_last"
  )

# write to csv for supp data
      #write.csv(summary_8h, file = "./suppData/8h_EPG-SummaryStatistics.csv")

summary_8h_full = LD_LL_8h %>%
  group_by(description, treatment, light) %>%
  summarise(
    max = max(value),
    min = min(value),
    n = n(),
    mean = mean(value),
    sd = sd(value),
    median = median(value),
    mode = DescTools::Mode(value)[1],
    .groups = "drop_last"
  )

# write to csv for supp data
      #write.csv(summary_8h_full, file = "./suppData/fullVariables_8h_EPG-SummaryStatistics.csv")

```

We can see that there are lots of zeros for the count data. We can also see the mode for the count data. For the duration data we can see that the only zeros are found in G and F which makes sense. We can also see that the modes for the duration data except for F are all NA, this is because most of these numbers are unique and there isn't any most frequent number.

With this knowledge we know that a gamma distribution will work for most of the duration variables except for F and G, in these cases we could use a hurdle model or better yet a tweedie model. For the count data, we can likely get by with a negative binomial distribution for all of our variables.

In E1 and E2, for LD - ZT0, we can see that there is at least one recording with a 0 value (min). I believe this was an aphid that remained dedicated in the pathway phase for a long duration of time. Since there was nothing technically wrong with this aphid, it was allowed to remain in the dataset since it is a true, observed behavior albeit an interesting/weird one. For these two behaviors we should maybe utilize a different model or add a small constant to the zeros.

Similar to the previous analysis. It's important to check and see if these models fit well. We can pick a few variables to analyze.

We should see how the fit is with the E1 and E2 duration when adding the constant and see if a different model (other than gamma glm) makes more sense.

Small sandbox to do some initial modeling

```{r}
# small modelling sandbox

# load required modeling and diagnostic packages
library(dplyr)
library(MASS)
# for diagnostics
library(DHARMa)
# type 3 anova
library(car)
# pairwise comps
library(emmeans)
# for the tweedie model to work with dharma use glmmTMB...
library(glmmTMB)

# prep the data 
sandbox_data = LD_LL_8h %>%
  filter(description %in% names(target_variables)) %>%
  mutate(value = as.numeric(value))

# since this is a dataset with more than two variables
# and we want to do type III anovas
# we want to set sum contrast rather than treatment contrasts
# this is since there isnt a clear baseline
options(contrasts = c("contr.sum", "contr.poly"))

# 1. gamma model (all durations > 0)
# "Total duration of C"
data_tot_dur_c = sandbox_data %>% filter(description == "Total duration of C")

# fit gamma with log link. including an interaction
m_gamma_durC = glm(value ~ treatment * light, family = Gamma(link = "log"),
              data = data_tot_dur_c)

# evaluate gamma Model
print(summary(m_gamma_durC), show.residuals = TRUE)

# base diagnostics
par(mfrow = c(2,2))
plot(m_gamma_durC, main = "gamma: Duration of C")

# reset plot window to default
par(mfrow = c(1,1))

# DHARMa 
res_gamma = simulateResiduals(fittedModel = m_gamma_durC, plot = TRUE)

# type III anova from car package
anova_durC = Anova(m_gamma_durC, type = "III")
print(anova_durC)

# pariwise comps with FDR adjustment
emmeans(m_gamma_durC, 
        pairwise ~ treatment:light, 
        type = "response", adjust = "fdr")

# calculate the estimated marginal means (EMMs) for the interaction
# type = "response" back-transforms the log link so estimates are in real -
# duration units
emm_durC = emmeans(m_gamma_durC, ~ treatment * light, type = "response",
                   adjust = "fdr")

# comp A: compare ZT treatments within each light condition
# this answers: "Is there a significant rhythm/difference across timepoints - 
# under LD? What about under LL?"
pairs_within_light = pairs(emm_durC, simple = "treatment", adjust = "fdr")
print(pairs_within_light)

# comp B: compare LD vs LL within each specific ZT timepoint
# this answers: "At ZT8 specifically, does the lighting condition change -
# the feeding duration?"
pairs_within_trt = pairs(emm_durC, simple = "light", adjust = "fdr")
print(pairs_within_trt)

# we can see in the Anova, that the interaction is the only thing showing any -
# p-value less than the significance threshold of 0.05.
# 
# further, when we do emmeans pairwise across all combos, with FDR correction,-
#nothing is significant but a few at ~ 0.08.
# 
# then, when we simply ask are there differences in means between the -
# individual treatments of light and time, we can see some significant -
#differences.
# 
# now lets try a NB, a tweedie, and see what models give us good fits for -
# the E1/E2 durations that only have a few zeros.

# cleanup
rm(res_gamma, pairs_within_light, 
   pairs_within_trt, m_gamma_durC, 
   emm_durC, data_tot_dur_c, anova_durC)



# 2. tweedie (durations with exact 0s)
# "Total duration of F"

data_tweedie_durF = sandbox_data %>% filter(description == "Total duration of F")

#
options(contrasts = c("contr.sum", "contr.poly"))

# fit tweedie model
m_tweedie_durF = glmmTMB(value ~ treatment * light, 
                        family = tweedie(link = "log"), 
                        data = data_tweedie_durF)

# eval tweedie Model
print(summary(m_tweedie_durF), show.residuals = TRUE)

# DHARMa
res_tweedie = simulateResiduals(fittedModel = m_tweedie_durF, plot = TRUE)

# ANOVA
anova_durF = Anova(m_tweedie_durF, type = "III")
print(anova_durF)

# no significant effects in treatment, light or the interaction for this -
# behavior. No need for pairwise comps.

# clean up
rm(anova_durF, data_tweedie_durF, m_tweedie_durF, res_tweedie)




# 3. neg binom (for 0 counts)
# "Number of C"

data_negbin_numC = sandbox_data %>% filter(description == "Number of C")

# fit negative binomial
m_negbin_numC = glm.nb(value ~ treatment*light, data = data_negbin_numC)

# eval nb model
print(summary(m_negbin_numC), show.residuals = TRUE)

# base diagnostics
par(mfrow = c(2,2))
plot(m_negbin_numC, main = "NegBin: Number of C")

# reset plot window to default
par(mfrow = c(1,1))

# DHARMa
res_negbin = simulateResiduals(fittedModel = m_negbin_numC, plot = TRUE)

# run anova
anova_numC = Anova(m_negbin_numC, type = "III")
print(anova_numC)

# pariwise comps with FDR adjustment
emmeans(m_negbin_numC, 
        pairwise ~ treatment:light, 
        type = "response", adjust = "fdr")

# calculate the estimated marginal means (EMMs) for the interaction
# type = "response" back-transforms the log link so estimates are in real -
# duration units
emm_numC = emmeans(m_negbin_numC, ~ treatment * light, type = "response",
                   adjust = "fdr")

# comp A: compare ZT treatments within each light condition
# this answers: "Is there a significant rhythm/difference across timepoints - 
# under LD? What about under LL?"
pairs_within_light_numC = pairs(emm_numC, simple = "treatment", adjust = "fdr")
print(pairs_within_light_numC)

# comp B: compare LD vs LL within each specific ZT timepoint
# this answers: "At ZT8 specifically, does the lighting condition change -
# the feeding duration?"
pairs_within_trt_numC = pairs(emm_numC, simple = "light", adjust = "fdr")
print(pairs_within_trt_numC)

# great we can see that everything worked well...
# now we can peek at the durE1 and E2 variables to see if gamma will work...

# clean up
rm(data_negbin_numC,
  m_negbin_numC,
  res_negbin,
  anova_numC,
  emm_numC,
  pairs_within_light_numC,
  pairs_within_trt_numC)



# 4. Starting with dur E1
# "Total duration of E1"
data_tot_dur_e1 = sandbox_data %>% 
  filter(description == "Total duration of E1")

# fit gamma with log link. including an interaction
# tweedie will push var closer to 2?
m_tweedie_durE1 = glmmTMB(value ~ treatment * light, 
                          family = tweedie(link = "log"), 
                          data = data_tot_dur_e1)

# summarize model
print(summary(m_tweedie_durE1), show.residuals = TRUE)


# dharma
res_tweedie_durE1 = simulateResiduals(fittedModel = m_tweedie_durE1, 
                                      plot = TRUE)

# anova
anova_durE1 = Anova(m_tweedie_durE1, type = "III")
print(anova_durE1)
# no sig differences in anova can move on to E2. Tweedie likely best.
# cleanup
rm(data_tot_dur_e1,
  m_tweedie_durE1,
  res_tweedie_durE1,
  anova_durE1)





# 5. Starting with dur E2
# "Total duration of E2"
data_tot_dur_E2 = sandbox_data %>% 
  filter(description == "Total duration of E2")

# fit gamma with log link. including an interaction
# tweedie will push var closer to 2?
m_tweedie_durE2 = glmmTMB(value ~ treatment * light, 
                          family = tweedie(link = "log"), 
                          data = data_tot_dur_E2)

# summarize model
print(summary(m_tweedie_durE2), show.residuals = TRUE)


# dharma
res_tweedie_durE2 = simulateResiduals(fittedModel = m_tweedie_durE2, 
                                      plot = TRUE)

# anova
anova_durE2 = Anova(m_tweedie_durE2, type = "III")
print(anova_durE2)

# we see a sig diff in the interaction. Lets run emmeans.

# pairwise comps with FDR adjustment
emmeans(m_tweedie_durE2, 
        pairwise ~ treatment:light, 
        type = "response", adjust = "fdr")

# calculate the estimated marginal means (EMMs) for the interaction
# type = "response" back-transforms the log link so estimates are in real -
# duration units
emm_durE2 = emmeans(m_tweedie_durE2, ~ treatment * light, type = "response",
                   adjust = "fdr")

# comp A: compare ZT treatments within each light condition
# this answers: "Is there a significant rhythm/difference across time points - 
# under LD? What about under LL?"
pairs_within_light_durE2 = pairs(emm_durE2, simple = "treatment", adjust = "fdr")
print(pairs_within_light_durE2)

# comp B: compare LD vs LL within each specific ZT time point
# this answers: "At ZT8 specifically, does the lighting condition change -
# the feeding duration?"
pairs_within_trt_durE2 = pairs(emm_durE2, simple = "light", adjust = "fdr")
print(pairs_within_trt_durE2)

# we have a failed KS test which is suggesting that the data doesn't perfectly -
# fit the distribution at hand. Everything else is good though. I think this -
# will be adequate. 

```

We can see that for this stuff the same applies as the 24 h data. Gamma for all the duration, except for F, G, E1, and E2 where we can do a Tweedie model due to zeros, and negative binomial for all of the counts!

# 2. Do GLM modelling and make plots.

Now we want these stats for all of our target variables and create plots that are overlaying the significant statistics onto the plot.

```{r}
# load libraries if not already
library(ggplot2)
library(dplyr)
# for models
# neg binom
library(MASS)
# type 3 anova
library(car)       
# tweedie models w/ anova + dharma support
library(glmmTMB)   
# pairwise comparisons
library(emmeans)   
# pretty plots for pub.
library(ggprism)

# make sure to set sum contrasts since we are using type III anovas...
options(contrasts = c("contr.sum", "contr.poly"))

# grab the data if haven't already
target_data = LD_LL_8h %>%
  filter(description %in% names(target_variables)) %>%
  mutate(value = as.numeric(value))

# make the empty list for the plots
boxplot_list_8h = list()

# empty list to store models 
model_list_8h = list()

# lists to store anova and pairwise comps
anova_list_8h = list() 
pairs_list_8h = list() 


# start looping through the EPG variable
for (var in names(target_variables)) {
  
  # grab units from target variables vector at each level of var
  unit = target_variables[[var]]
  # grab data for each level of var
  plot_data = target_data %>% filter(description == var)
  
  # start unit conversion logic
  if (unit == "h") {
    plot_data$value = plot_data$value / 3600
    y_label = paste(var, "(hr)")
  } else if (unit == "min") {
    plot_data$value = plot_data$value / 60
    y_label = paste(var, "(min)")
  } else if (unit == "sec") {
    y_label = paste(var, "(sec)")
  } else {
    y_label = paste(var)
  }
  
  # build the base plot 
  p = ggplot(plot_data, aes(x = treatment, y = value, fill = light)) +
    geom_violin(position = position_dodge(1), alpha = 1, color = NA) + 
    geom_boxplot(width = 0.5, color = "grey10", outlier.shape = NA, 
                 position = position_dodge(1), alpha = 0.5) +
    geom_point(aes(shape = light), 
               position = position_jitterdodge(dodge.width = 1, jitter.width = 0.5),
               color = "black", alpha = 0.5) +
    scale_shape_manual(
      values = c("LD" = 21, "LL" = 22) 
    ) +
    scale_fill_manual(values = c("LD" = "#0077BB", "LL" = "#EE7733")) +
    scale_x_discrete(labels = c("ZT0" = "0", "ZT8" = "8", "ZT16" = "16")) +
    guides(fill = guide_legend(position = "inside"),
             shape = guide_legend(position = "inside")) +
    theme_prism() +
    theme(legend.position.inside = c(.95, .95)) +
    labs(x = "Zeitgeber / Circadian Time (h)", y = y_label, fill = "Light Regime", shape = "Light Regime")
  
  tryCatch({
    # 1. neg binom for all count data
    if (unit == "count") {
      mod = glm.nb(value ~ treatment * light, data = plot_data)
      
    # 2. glmmTMB Tweedie for specific zero-inflated durations
    } else if (var %in% c("Total duration of F", "Duration of G", 
                          "Total duration of E1", "Total duration of E2")) {
      mod = glmmTMB(value ~ treatment * light, family = tweedie(link = "log"), 
                    data = plot_data)
      
    # 3. gamma for all other duration vars
    } else {
      mod = glm(value ~ treatment * light, family = Gamma(link = "log"), 
                data = plot_data)
    }
    
    # save fitted model to empty list with var as name
    model_list_8h[[var]] = mod
    
    # run type 3 anova to check global significance
    aov_res = Anova(mod, type = "III")
    
    # save anova to list
    anova_list_8h[[var]] = aov_res
    
    # grab anova cols
    p_col = grep("Pr", colnames(aov_res), value = TRUE)[1]
    
    p_trt = ifelse("treatment" %in% rownames(aov_res), 
                   aov_res["treatment", p_col], NA)
    p_lgt = ifelse("light" %in% rownames(aov_res), 
                   aov_res["light", p_col], NA)
    p_int = ifelse("treatment:light" %in% rownames(aov_res), 
                   aov_res["treatment:light", p_col], NA)
    
    # only proceed with post-hoc brackets if ANY main effect or interaction is significant
    if (any(c(p_trt, p_lgt, p_int) < 0.05, na.rm = TRUE)) {
      
      # run emmeans for the interaction
      emm = emmeans(mod, ~ treatment * light, type = "response", 
                    adjust = "fdr")
      
      
      # extract raw pairwise dataframes to save to list
      raw_pairs_lgt = as.data.frame(pairs(emm, simple = "light", adjust = "fdr"))
      raw_pairs_trt = as.data.frame(pairs(emm, simple = "treatment", adjust = "fdr"))
      
      # save raw pairwise comps to list, significant or not, all saved
      pairs_list_8h[[var]] = bind_rows(
        raw_pairs_lgt %>% mutate(comp_type = "Within-ZT (LD vs LL)"),
        raw_pairs_trt %>% mutate(comp_type = "Across-ZT (Temporal Changes)")
      )
      
      # extract filtered pairwise dataframes for plotting, only plot sig
      pairs_lgt = as.data.frame(pairs(emm, simple = "light", 
                                      adjust = "fdr")) %>% 
        filter(p.value < 0.05)
      
      pairs_trt = as.data.frame(pairs(emm, simple = "treatment", 
                                      adjust = "fdr")) %>% 
        filter(p.value < 0.05)
      
      # initialize dynamic Y tracker for staggering brackets across comps
      max_y = max(plot_data$value, na.rm = TRUE)
      current_y = max_y + (max_y * 0.05)
      step_y = max_y * 0.08
      
      #  plotting Comparisons within ZT (LD vs LL) 
      if (nrow(pairs_lgt) > 0) {
        for (i in 1:nrow(pairs_lgt)) {
          row = pairs_lgt[i, ]
          
          # find X coordinates (dodged offset ~ 0.2 around integer center)
          x_center = match(as.character(row$treatment), 
                           levels(plot_data$treatment))
          x1 = x_center - 0.2  # LD
          x2 = x_center + 0.2  # LL
          
          # set asterisks
          sig_stars = case_when(row$p.value < 0.001 ~ "***",
                                row$p.value < 0.01  ~ "**",
                                row$p.value < 0.05  ~ "*")
          # annotate plot with the brackets and text
          p = p + 
            annotate("segment", x = x1, xend = x2, y = current_y, 
                     yend = current_y, linewidth = 0.75) +
            annotate("text", x = x_center, y = current_y + (max_y * 0.02), 
                     label = sig_stars, size = 5, fontface = "plain")
          
          current_y = current_y + step_y
        }
      }
      
      # plotting comparisons within light treat (e.g. ZT0 vs ZT8)
      if (nrow(pairs_trt) > 0) {
        for (i in 1:nrow(pairs_trt)) {
          row = pairs_trt[i, ]
          
          # set X offset based on light regime
          offset = ifelse(as.character(row$light) == "LD", -0.2, 0.2)
          
          # parse contrast string (e.g., "ZT0 - ZT8" or "ZT0 / ZT8" depending on link function)
          zts = unlist(strsplit(as.character(row$contrast), "\\s+[-/]\\s+"))
          zts = gsub("[()]", "", zts) # clean parens if emmeans added them
          
          x1 = match(trimws(zts[1]), levels(plot_data$treatment)) + offset
          x2 = match(trimws(zts[2]), levels(plot_data$treatment)) + offset
          
          sig_stars = case_when(row$p.value < 0.001 ~ "***",
                                row$p.value < 0.01  ~ "**",
                                row$p.value < 0.05  ~ "*")
          # add annotations
          p = p + 
            annotate("segment", x = x1, xend = x2, y = current_y, 
                     yend = current_y, linewidth = 0.75) +
            annotate("text", x = mean(c(x1, x2)), 
                     y = current_y + (max_y * 0.02), 
                     label = sig_stars, size = 5, fontface = "plain")
          
          current_y = current_y + step_y
        }
      }
      
      # final coordinate expansion based on all drawn brackets
      p = p + coord_cartesian(ylim = c(0, current_y + (max_y * 0.05)))
    }
    # debug
  }, error = function(e) {
    p <<- p + labs(subtitle = "Model failed to converge/evaluate")
  })
  
  # store the plot to list and then print
  boxplot_list_8h[[var]] = p
  print(p)
}
```

Save to svg looping through list...

```{r}
# global width and height
plot_width = 5
plot_height = 5


# loop through the list and save each plot
for (var_name in names(boxplot_list_8h)) {
  
  # clean up the name for the file system
  safe_name = gsub(" ", "_", var_name)
  file_path = paste0("./figures_8h/boxplots/", safe_name, "upd_x-axis_.svg")
  
  # save as svg
  # ggsave(
  #   filename = file_path,
  #   plot = boxplot_list_8h[[var_name]],
  #   width = plot_width, 
  #   height = plot_height,
  #   device = "svg"
  # )
  
  print(paste("Saved plot:", file_path))
}
```

Now we can create some multi-plots that contain duration of C, duration of E1, duration of E2, number of E1.

These should be pub. ready.

Manually dropping X axis, i.e. collecting to bottom row of plots.

```{r}
library(patchwork)

# define grouping vars
specific_vars = c("Total duration of C", "Total duration of E2", 
                  "Number of np", "Number of pd",
                  "Number of C","Number of E1")


# debug
available_vars = intersect(specific_vars, names(boxplot_list_8h))


# start
if (length(available_vars) > 0) {
  
  # extract the plots into a sublist
  multi_plots = boxplot_list_8h[available_vars]
  
  # strip the x-axis title from all plots EXCEPT the bottom row
  # for a 6-plot, 2-column grid, this removes the label from plots 1 through 4
  for (i in 1:(length(multi_plots) - 2)) {
    multi_plots[[i]] = multi_plots[[i]] + theme(axis.title.x = element_blank())
  }
  
  # combine using patchwork syntax
  combined_plot_1 = wrap_plots(multi_plots, ncol = 2) + 
    # dropped axis_titles = "collect_x" since handling it manually above
    plot_layout(guides = "collect") &  
    theme(
      legend.position = "none"
    ) &      
    guides(fill = guide_legend(nrow = 1, byrow = TRUE),
      shape = guide_legend(nrow = 1, byrow = TRUE)) &
    plot_annotation(tag_levels = 'A')
  
  # ggsave(
  #   filename = "./figures_8h/multiplots/variables_durations_and_counts_8h_nolegend_neworder_axisv2.svg",
  #   plot = combined_plot_1,
  #   width = 7.5,
  #   height = 10,
  #   device = "svg"
  # )
  print(combined_plot_1)
  print("Saved Multiplot 1 successfully!")
}
```

We want to automate the output of these tables onto word documents to easily make supplementary tables for submission 

we can do that in this code chunk

```{r}
library(dplyr)
library(purrr)
library(broom)
library(broom.mixed)
library(car)
library(flextable) 
library(officer) 
library(emmeans)

# define the slightly-cleaned formatting function
format_model_to_flextable_clean_names = function(model, var_name) {
  
  # extract and rename coeffs
  tidy_coef = tryCatch({
    broom.mixed::tidy(model, effects = "fixed")
  }, error = function(e) {
    broom::tidy(model)
  })
  
  # swap terms in the term column to names that match manuscript
  tidy_coef = tidy_coef %>%
    mutate(
      term = gsub("treatment", "Time of Day ", term),
      term = gsub("light", "Light Regime ", term),
      term = gsub(":", " \u00D7 ", term) # Also makes the interaction symbol look professional
    )
  
  # extract and rename anova
  anova_test = tryCatch({
    car::Anova(model, type = "III") %>% 
      broom::tidy() %>% 
      filter(term != "(Intercept)") 
  }, error = function(e) {
    NULL
  })
  
  if(!is.null(anova_test)) {
    anova_test = anova_test %>%
      mutate(
        term = gsub("treatment", "Time of Day", term),
        term = gsub("light", "Light Regime", term),
        term = gsub(":", " \u00D7 ", term)
      )
  }
  
  # if sig, extract and rename pairwise results
  pairs_lgt_data = NULL
  pairs_trt_data = NULL
  
  if(!is.null(anova_test) && any(anova_test$p.value < 0.05, na.rm = TRUE)) {
    tryCatch({
      emm = emmeans(model, ~ treatment * light, type = "response", adjust = "fdr")
      
      # extract and rename the 'treatment' column to 'Time of Day'
      pairs_lgt_data = as.data.frame(pairs(emm, simple = "light", adjust = "fdr")) %>%
        rename(`Time of Day` = treatment) 
      
      # extract and rename the 'light' column to 'Light Regime'
      pairs_trt_data = as.data.frame(pairs(emm, simple = "treatment", adjust = "fdr")) %>%
        rename(`Light Regime` = light)    
      
    }, error = function(e) {
      message(paste("Note: Pairwise extraction skipped or failed for", var_name, "-", e$message))
    })
  }
  
  # build the flextables
  # adding a small helper to add titles above tables
  make_basic_ft = function(df) {
    if(is.null(df)) return(NULL)
    
    flextable(df) %>%
      colformat_double(digits = 2) %>% 
      theme_booktabs() %>% 
      autofit()
  }
  
  # apply helper
  coef_ft      = make_basic_ft(tidy_coef)
  anova_ft     = make_basic_ft(anova_test)
  pairs_lgt_ft = make_basic_ft(pairs_lgt_data)
  pairs_trt_ft = make_basic_ft(pairs_trt_data)
  
  return(list(coefs = coef_ft, anova = anova_ft, pairs_lgt = pairs_lgt_ft, pairs_trt = pairs_trt_ft))
}


# get directory and start loop to make files
dest_dir = file.path(getwd(), "tables_8h")
if(!dir.exists(dest_dir)) dir.create(dest_dir)

for (var_idx in names(model_list_8h)) {
  
  safe_name = gsub(" ", "_", var_idx)
  tables = format_model_to_flextable_clean_names(model_list_8h[[var_idx]], var_idx)
  file_path = file.path(dest_dir, paste0(safe_name, "_8h_Results.docx"))
  
  # start blank word doc.
  doc = read_docx()
  
  # add files sequentially
  if(!is.null(tables$anova)) {
    doc = body_add_par(doc, paste("Type III ANOVA:", var_idx), style = "Normal")
    doc = body_add_flextable(doc, value = tables$anova)
    doc = body_add_par(doc, "") 
  }
  
  if(!is.null(tables$coefs)) {
    doc = body_add_par(doc, paste("Model Coefficients:", var_idx), style = "Normal")
    doc = body_add_flextable(doc, value = tables$coefs)
    doc = body_add_par(doc, "")
  }
  
  if(!is.null(tables$pairs_lgt)) {
    doc = body_add_par(doc, paste("Pairwise (LD vs LL within ZT):", var_idx), style = "Normal")
    doc = body_add_flextable(doc, value = tables$pairs_lgt)
    doc = body_add_par(doc, "")
  }
  
  if(!is.null(tables$pairs_trt)) {
    doc = body_add_par(doc, paste("Pairwise (Temporal changes within Light Regime):", var_idx), style = "Normal")
    doc = body_add_flextable(doc, value = tables$pairs_trt)
  }
  
  # save the fully built document
  print(doc, target = file_path)
  
  print(paste("Successfully saved Word document for:", var_idx))
}
```



# 3. EPG Summary Plots

We can create the EPG summary plots. We want to create multi-panel plots. Ideally we want these faceted by LD and LL rows with columns for treatment (ZT time). We can try the other way to check if it looks better.

Below we can grab the data and format it accordingly.

```{r}
# discoEPG and patchwork
library(discoEPG)
library(patchwork)
library(ggprism)
# labels
waveform_labels = c("np" = 1,
                    "C" = 2,
                    "E1e" = 3,
                    "E1" = 4,
                    "E2" = 5,
                    "F" = 6,
                    "G" = 7,
                    "pd" = 8,
                    "end" = 12)


# pass EPG dir for ana
ana_dir_LD = "../EPG_data/8h_data/LD/LD_all_ANA/"
ana_dir_LL = "../EPG_data/8h_data/LL/LL_all_ANA/"

# now we can grab ana file names. Remember 1-4 LD and 5-8 LL
filenames_LD = list.files(path = ana_dir_LD, pattern = ".ANA")
filenames_LL = list.files(path = ana_dir_LL, pattern = ".ANA")

# load LD data
anaDatList_LD = discoEPG::combine_ana(directory = ana_dir_LD,
                                     filenames = filenames_LD,
                                     validate = TRUE,
                                     waveform_labels = waveform_labels)
# extract data
anaData_LD = anaDatList_LD$data

# add treatment
anaData_LD$treatment = "LD" 

# load LL data
anaDatList_LL = discoEPG::combine_ana(directory = ana_dir_LL,
                                     filenames = filenames_LL,
                                     validate = TRUE,
                                     waveform_labels = waveform_labels)
# extract data
anaData_LL = anaDatList_LL$data

# add treatment
anaData_LL$treatment = "LL"

# combine them into a single master data frame
anaData_master = rbind(anaData_LD, anaData_LL)

### assign light treatments

# grab all unique filenames across both datasets
allfiles = unique(anaData_master$filename)

# define file vectors based on zt strings found in the filenames
zt0_files  = grep("0zt",  allfiles, value = TRUE, ignore.case = TRUE)
zt8_files  = grep("8zt",  allfiles, value = TRUE, ignore.case = TRUE)
zt16_files = grep("16zt", allfiles, value = TRUE, ignore.case = TRUE)

# assign the ZT factor column, be sure to factorize...
anaData_master$zt = NULL
anaData_master$zt[anaData_master$filename %in% zt0_files]  = "ZT0"
anaData_master$zt[anaData_master$filename %in% zt8_files]  = "ZT8"
anaData_master$zt[anaData_master$filename %in% zt16_files] = "ZT16"

# check if any files missed assignment
if(any(is.na(anaData_master$zt))) {
  warning("Some files were not assigned a ZT group! Check your filename patterns.")
}
```

Now start creating the plots.

```{r}
# define combinations and colors
treatments = c("LD", "LL")
zt_groups = c("ZT0", "ZT8", "ZT16")

# set color palette colorblind safe (paul tol)
behavior_colors = c(
  "NP" = "black", "C"  = "#0077BB", "E1" = "#CC3311",
  "E2" = "#009988", "F"  = "#EE7733", "G"  = "#33BBEE", "PD" = "#EE3377"
)
```

Facet plots below.

```{r}
library(tidyverse)

# combind prop data
combined_prop_data = anaData_master %>%
  group_split(treatment, zt) %>%
  map_dfr(function(df) {
    tr = unique(df$treatment)
    z = unique(df$zt)
    
    # grab data from combined 
    res = discoEPG::ana_to_timeseries_proportion(df, waveform_labels)
    
    # take the time data and add +8 to convert to ZT time
    # as each recording starts at zt and is 8h long
    res$data %>%
      mutate(
        treatment = tr,
        zt = factor(z, levels = c("ZT0", "ZT8", "ZT16")),
        # shift the timeline based on the ZT group
        hours = case_when(
          zt == "ZT0"  ~ hours,
          zt == "ZT8"  ~ hours + 8,
          zt == "ZT16" ~ hours + 16,
          TRUE ~ hours
        )
      )
  })

# define the shading box for hour 16-24
shade_df = expand.grid(
  treatment = c("LD", "LL"), 
  zt = factor("ZT16", levels = c("ZT0", "ZT8", "ZT16"))
) %>%
  mutate(
    xmin = 16, xmax = 24, 
    ymin = -Inf, ymax = Inf
  )

# start plot 
plot_LD = ggplot(combined_prop_data %>% filter(treatment == "LD"), 
                              aes(x = hours, 
                                  y = proportion, 
                                  color = activity)
                              ) +
  # targeted shading layer mapped strictly to ZT16
  geom_rect(
    data = shade_df %>% filter(treatment == "LD"),
    aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = ymax),
    inherit.aes = FALSE,
    fill = "grey50", alpha = 0.15
  ) +
  
# smooth line data
  # stat_smooth(method = "gam",
  #             formula = y ~ s(x, bs = "cs", k = 15),
  #             se = F, linewidth = 1, alpha = 0.2
  #             ) +
  
  stat_smooth(method = "gam", 
              formula = y ~ s(x, bs = "cs", k = 40), 
              se = F, linewidth = 1, alpha = 0.2) +

  # geom_line(linewidth = 1) +
  
  # grid with free x for zeitgeber time 
  facet_grid(. ~ zt, scales = "free_x") + 
  
  # themes and explicit y axis breaks
  theme_prism() +
  scale_y_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1), limits = c(0, 1)) +
  
  # ensure the breaks fall cleanly on even numbers for all scales
  scale_x_continuous(breaks = scales::breaks_width(2)) + 
  
  # new colors tol
  scale_color_manual(
    values = scales::alpha(behavior_colors, 0.8),
    breaks = c('NP', 'C', 'PD', 'E1', 'E2', 'F', 'G')
  ) +
  # labs 
  labs(x = "Zeitgeber Time (h)", y = "Proportion of Aphids", color = "Behavior") +
  # upd theme with new info. add panel spacing between facets
  theme(
    legend.position = "right",
    panel.spacing.y = unit(2, "lines"),
    panel.spacing.x = unit(1, "lines")
    #legend.title = element_text("Behavior")
  ) +
  # guide line width thicker
  guides(color = guide_legend(override.aes = list(linewidth = 2)))


plot_LL = ggplot(combined_prop_data %>% filter(treatment == "LL") %>% mutate(ct_label = factor(str_replace(zt, "ZT", "CT"), levels = c("CT0", "CT8", "CT16"))), 
                              aes(x = hours, 
                                  y = proportion, 
                                  color = activity)
                              ) +
  
# smooth line data
  # stat_smooth(method = "gam",
  #             formula = y ~ s(x, bs = "cs", k = 15),
  #             se = F, linewidth = 1, alpha = 0.2
  #             ) +
  
  stat_smooth(method = "gam", 
              formula = y ~ s(x, bs = "cs", k = 40), 
              se = F, linewidth = 1, alpha = 0.2) +

  # geom_line(linewidth = 1) +
  
  # grid with free x for circadian time 
  facet_grid(. ~ ct_label, scales = "free_x") + 
  
  # themes and explicit y axis breaks
  theme_prism() +
  scale_y_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1), limits = c(0, 1)) +
  
  # ensure the breaks fall cleanly on even numbers for all scales
  scale_x_continuous(breaks = scales::breaks_width(2)) + 
  
  # new colors tol
  scale_color_manual(
    values = scales::alpha(behavior_colors, 0.8),
    breaks = c('NP', 'C', 'PD', 'E1', 'E2', 'F', 'G')
  ) +
  # labs 
  labs(x = "Circadian Time (h)", y = "Proportion of Aphids", color = "Behavior") +
  # upd theme with new info. add panel spacing between facets
  theme(
    legend.position = "right",
    panel.spacing.y = unit(2, "lines"),
    panel.spacing.x = unit(1, "lines")
    #legend.title = element_text("Behavior")
  ) +
  # guide line width thicker
  guides(color = guide_legend(override.aes = list(linewidth = 2)))

# combine via patchwork
faceted_summary_plot = plot_LD / plot_LL + plot_layout(guides = "collect")

# print. sit tight it takes a while due to smoothing
print(faceted_summary_plot)
```

Save the plots to svg/pdf

```{r}
# saving the combined summary plot
# sit tight, takes a minute. large plot...
    # ggsave(
    #   filename = "./figures_8h/summary/TimeSeries_Summary_8h_newcolors_upd_facet.svg",
    #   plot = faceted_summary_plot,
    #   width = 7.5,
    #   height = 6,
    #   device = "svg"
    # )
```
