---
title: "LD vs LL Analysis (8h and 24h EPG recordings)"
author: "Daniel Kunk"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
---

# LD-LL and 24h EPG Analysis

This document will analyze EPG data that was collected for 8 hour duration over three time points (ZT/CT0, 8 and 16) in both LD and LL conditions (16:8 light:dark or 24h light). Additionally, we will analyze data that was collected at one time point continuously for 24 hours between LD and LL conditions.

### Install any necessary packages.

First, we will install any necessary packages. These will be stored in our R-project virtual environment. This means that someone with this repository/R-project should be able to re-produce our analysis exactly. First we will install the tidyverse which should handle most of our needs. In addition we can install the EPGplotR package which will allow us to create some beautiful plots of our EPG signals.

Remember, you may not need to do this, since I am installing these in an R-project R virtual environment. As long as you have this repository you should be good. You can always look in your console for the `Project 'C:/Users/danie/OneDrive - Colostate/NALAM LAB/Authored Papers/DK_Aphid_Feeding_2026/R_scripts/LDvsLL_24h_EPG_2026' loaded. [renv 1.1.5]` line showing that you are in the right place and the renv is loaded up. Just load a package with `library()` and if it loads you should be all good!

```{r}
# We only need a few packages.

# install tidyverse
### install.packages("tidyverse")


# now to install discoEPG we need to install some more stuff!
# devtools and pak package allows us to install from source code on github.
# (devtools isnt required feel free to remove here)
# It installs missing one dependency so we can install that after.
### install.packages("pak")
### install.packages("gtools")
#library(pak)

# use pak to install from source, make sure you have rtools installed (rtools45)
### pak("dankunk/EPGplotR")

# check to see if EPGplotR is installed.
library(EPGplotR)

# Now lets install some packages for non-parametric two-way anova
### install.packages("ARTool")

# some additional packages useful for modelling in R
### install.packages("tidyr")
### install.packages("emmeans")
### install.packages("ggsignif")


# now we need to install the discoEPG R package source to be able to make timeseries mode (i.e. most frequent) plots of each waveform per time step.

### pak("nalamvj/EPG_ANALYSIS/discoEPG")

# Check to see if installed...
### library(discoEPG)

```

### Begin the Analysis

Now that we have some packages installed we can start to do our analysis. First lets plan out what we want to do in steps.

Analyze LD and LL - 8h EPG data:

1.  Perform statistics between time points (within light-treatment) and between light-treatment within time points and generate formatted/documented output tables for suitable for publication.

2.  Generate figures with statistics over each bar/box/etc. for the comparisons in question for the major EPG variables.

3.  Generate summary figures with discoEPG plots and GLM plots to summarize behavior over the feeding windows.

Analyze LD and LL - 24h EPG data:

1.  Perform Statistics between LD and LL recordings.

2.  Generate figures with statistics for the comparisons in question for major EPG variables.

3.  Generate figures of the 24h signal for each recording and fit into a large multi-panel plot (1 column, n rows) that will be contained in the supplementary figures.

Note: This data was collected using the GIGA8dd system. Please see materials and methods for full entrainment and recording protocol. Further, the raw data was analyzed using the Stylet+ ANA program. ANA files were imported into the discoEPG shiny app for EPG variable calculation (v1.0; now deprecated see discoEPG python package instead...).

## Lets start with the 8h LD and LL data analysis.

### A. To start lets load, format, and explore the data:

```{r}
library(tidyverse)

# load the data and add a light treatment column

LD_8h = read.csv("./data/LD_ShinyAppParameterDataExport_2026-04-07 20_03_26.csv") %>% mutate(light = "LD")

LL_8h = read.csv("./data/LL_ShinyAppParameterDataExport_2026-04-07 20_06_57.csv") %>% mutate(light = "LL")


# Join the two data frames on top of one another
LD_LL_8h = rbind(LD_8h, LL_8h)

# we have some inf and NA values that we will drop for now.
LD_LL_8h = LD_LL_8h %>%
  filter(is.finite(value))

# the natural tendency of R is to sort by first number so ZT16 comes before 8.
# Our treatment variable (ZT) has a natural order so lets set that.
LD_LL_8h$treatment = factor(LD_LL_8h$treatment, levels = c("ZT0", "ZT8", "ZT16"), ordered = T)

# Downstream modeling likes our variables to be factors so we should set light as well
LD_LL_8h$light = factor(LD_LL_8h$light, levels = c("LD","LL"), ordered = F)


```

Now some slight EDA before hypothesis testing:

```{r}
# if you are interested in the dataset we can see:
summary(LD_LL_8h)

# file name: the epg recording analysis file
# value: the value of the variable
# description: semantic name of variable
# acronym: acronym of variable
# treatment: time of recording start (ZT)
# light: light regime (LD or LL)

# furthermore glimpse the data
head(LD_LL_8h)

# now lets check each the mean and std dev of every variable across all ind.
LD_LL_8h %>%
  group_by(description) %>%
  # Now you can safely run means or hypothesis tests
    summarise(
      mean = mean(value),
      stddev = sd(value),
      min = min(value),
      max = max(value),
      q25 = quantile(value, 0.25),
      median = quantile(value, 0.50), # or median(value)
      q75 = quantile(value, 0.75),
      n = n(),
      .groups = "drop"
    )

# We can also check the mean and std per treatments by grouping!
LD_LL_8h %>%
  group_by(description, treatment, light) %>%
  # Now you can safely run means or hypothesis tests
    summarise(
      mean = mean(value),
      stddev = sd(value),
      min = min(value),
      max = max(value),
      q25 = quantile(value, 0.25),
      median = quantile(value, 0.50), # or median(value)
      q75 = quantile(value, 0.75),
      n = n(),
      .groups = "drop"
    )
```

-   Now that we have an idea about the centrality and variance of our data, we should talk about our hypotheses. Because this is a general or exploratory experiment our hypothesis is rather general.

$$H_0: \text{The means of behaviors are identical across Light and ZT groups.}$$

$$and$$

$$H_A: \text{At least one group mean differs significantly.}$$

-   Usually, a two-way anova would do the trick here but lets check out our data:

```{r}
# say for this variable, Duration of C, how is the data distributed across all individuals
LD_LL_8h %>%
  filter(description == "Total duration of C") %>%
  ggplot(aes(x = value, fill = treatment)) + 
  geom_histogram(
    alpha = 0.25,
    bins = 20,
    color = "black"
  ) +
  labs(x = "duration (s)") +
  facet_wrap(. ~ light, nrow = 3)

# what if its transformed, does it become normal? 
LD_LL_8h %>%
  filter(description == "Total duration of C") %>%
  ggplot(aes(x = log10(value), fill = treatment)) + 
  geom_histogram(
    alpha = 0.25,
    bins = 20,
    color = "black"
  ) +
  labs(x = "Log10 duration (s)") +
  facet_wrap(. ~ light, nrow = 3)
# looks better but even if something is normal, it still might not meet the requirements for a parametric analysis.

```

-   We can see that our data doesn't even fit a normal distribution, it is right-Skewed. Further, we could do more tests to see if it was parametric (qq-plot, variance, shaprio test, etc.), but from experience I know that this data will mostly be non-parametric as it has never been for EPG data...

-   This means that we need to use a non-parametric two-way anova. Also known as a Aligned Rank Transformed two-way anova (ART).

### B. Now we can do our hypothesis testing:

Here we will use the ARTool package in addition to emmeans.

```{r echo=TRUE}
# load aligned rank transform package
library(ARTool)
library(emmeans)
library(kableExtra)

# grab temmeans# grab test df of 1 variable.
df_durC = LD_LL_8h %>% 
  filter(description == "Total duration of C")

# model the duration value as a function of treatment, light, and their interaction.
m.art = art(value ~ treatment*light, data = df_durC)

# peek at the model
m.art

# analysis of variance
anova(m.art)

# conducting contrasts
art.con(m.art, "treatment")
art.con(m.art, "light")

# contrasts
art.con(m.art, "treatment:light")

# we can pipe these dfs into kable for nice tex tables of the model data

anova_results = anova(m.art)

# Output as a LaTeX table
kable(anova_results, booktabs = TRUE, 
      caption = "ART ANOVA results for Total Duration of C") %>%
  kable_styling(latex_options = c("striped", "hold_position"))

```

### C. Now we can do some plotting:

```{r}

ggplot(df_durC, aes(x = treatment, y = value, fill = light)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(pch = 24, position = position_jitterdodge()) +
  scale_y_continuous(n.breaks = 8) +
  xlab("Timepoint (Zeitgeber time)") +
  ylab("Duration (s)") +
  theme_minimal() +
  coord_cartesian()

```

### D. Purring across relevant descriptions:

First, we can set a vector of relevant variables that we want to model.

```{r}
# vector of relevant variables.
# the more we include without p-correction the greater likelihood of finding something significant just by chance.

target_variables = c(
  "Total duration of np",
  "Total duration of pd",
  "Total duration of C",
  "Total duration of E1",
  "Total duration of E2",
  "Total duration of F",
  "Duration of G",
  "Number of np",
  "Number of pd",
  "Number of C",
  "Number of E1",
  "Number of E2",
  "Number of F",
  "Number of G"
)
```

### E. Next, we can start the pipeline mapping...

```{r}
# build the nested pipeline

art_pipeline = LD_LL_8h %>%
  # Filter the dataset down to ONLY your pre-selected list
  filter(description %in% target_variables) %>%
  
  # Group by your remaining EPG variables
  group_by(description) %>%
  nest() %>%
  
  # Run the ART model and ANOVA
  mutate(
    art_model = map(data, ~ art(value ~ treatment * light, data = .x)),
    anova_res = map(art_model, anova),
    anova_df = map(anova_res, ~ as_tibble(.x, rownames = "term"))
  )


```

################################################ 

# Starting the 24h analysis

Now we can start the 24h analysis.

-   First we need to load our data.

-   Then we can perform analysis between treatments.

-   Next we want to model the data across 24h using GLM on the mode of each data.

1.  First we can load the data.

```{r}
# load dataset
LD_LL_24 = read.csv("./data/24h_ShinyAppParameterDataExport_2026-04-07 19_48_18.csv")

# note that this dataset has a slightly different structure than the other two 8h datsets
```

############################################################## 

THIS IS THE timeseries modelling section. Use space above for the LD vs LL hypothesis testing analysis.

number X. Next we can prepare the data for some analysis

```{r}
# First we need to point to the ANA directories
# in above directory from our r project dir.
ana_dir_24 = "../EPG_data/24h_data/24h_ana/"

# now we can grab ana file names. Remember 1-4 LD and 5-8 LL
filenames = list.files(path = ana_dir_24, pattern = ".ANA")

# before we can load the data with discoEPG to perform mode calculations, we need to provide a list of waveform labels in a vector.
waveform_labels = c("np" = 1,
                    "C" = 2,
                    "E1e" = 3,
                    "E1" = 4,
                    "E2" = 5,
                    "F" = 6,
                    "G" = 7,
                    "pd" = 8,
                    "end" = 12)

# Now we can load the names into a list containing the data and filenames, remember this is a discoEPG R package function.
anaDatList_24 = combine_ana(directory = ana_dir_24,
                          filenames = filenames,
                          validate = TRUE, # perform data validation
                          waveform_labels = waveform_labels)

# extract the "data" dataframe from this list
anaData_24 = anaDatList_24$data

# We can also extract and print validation errors, this is a bit redundant since we have already done QC on the shiny app during variable export!
#cat(anaDatList$errors)

# grab all unique filenames
allfiles = unique(anaData_24$filename)

# assign treatments for files with names containing the string of interest
LD_treat_files = grep("ch[1-4]", allfiles, value = TRUE)
LL_treat_files = grep("ch[5-8]", allfiles, value = TRUE)

# set an empty variable for treatment to assign below.
anaData_24$treatment = NULL

# now we can set the row values for that column based on the presence in each xx_treat_files variable
anaData_24$treatment[anaData_24$filename %in% LD_treat_files] = "LD"
anaData_24$treatment[anaData_24$filename %in% LL_treat_files] = "LL"

# now since ana_to_timeseries can only work on one dataset at a time we can subset the dataset with the treatments we just assigned.
LD_24_ana_data = subset(anaData_24, treatment == "LD")
LL_24_ana_data = subset(anaData_24, treatment == "LL")

# now we can generate the modes for the timeseries data... This is a discoEPG function...
LD_24_modes = ana_to_timeseries(
  ana_df = LD_24_ana_data,
  waveform_labels = waveform_labels
)

LL_24_modes = ana_to_timeseries(
  ana_df = LL_24_ana_data,
  waveform_labels = waveform_labels
)

# this is great. We can generate plots with this data. But to go forward and model it, we want to format it a bit further.
LD_24_modes_dat = LD_24_modes$data
LL_24_modes_dat = LD_24_modes$data

# add treatment column again...
LD_24_modes_dat$treatment = "LD"
LL_24_modes_dat$treatment = "LL"

# combined dataframe...
LD_LL_24_modes_dat = rbind(LD_24_modes_dat, LL_24_modes_dat)
# note that this data is agnostic to the filename at this state int time. Might make sense to edit the discoEPG function to preserve a file name column in the dataframe output from ana_to_timeseries().

# in the next code chunk we can take this data and model it using a GLM approach to summarize the data...

```
