---
title: "Performance Assay Statistical Analysis"
author: "Daniel Kunk"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
---
# Performance Assay Analysis

Load and format the data if necessary

```{r}
library(dplyr)
# perf data above project dir
# total # column has '#' character, rename
perf_data = read.csv("../Perf_data/perf_data.csv") %>%
  rename(total_aphids = Total..)

```

Summarize the data and plot hist across all vars.

```{r}
# load packages
library(dplyr)
library(ggplot2)
library(ggpubr)
library(ggridges)

# summary stats
perf_sum = perf_data %>%
  group_by(treat) %>%
  summarise(
    mean_tot_adult = mean(adult),
    mean_tot_nymph = mean(nymph),
    mean_norm_nymph = mean(nym_per_aph),
    median_tot_adult = median(adult),
    median_tot_nymph = median(nymph),
    median_norm_nymph = median(nym_per_aph),
    sd_tot_nymph = sd(nymph), 
    sd_norm_nymph = sd(nym_per_aph),
    n = n()
  )

# save summary stats for suppData
    #write.csv(perf_sum, file = "./suppData/performance_SummaryStatistics.csv")

perf_data %>% ggplot(aes(x = nym_per_aph, y = factor(treat), fill = treat)) + 
  geom_density_ridges() + theme_ridges(center_axis_labels = T) + 
  xlab("Nymphs per Adult Aphid") + ylab(NULL) + theme(legend.position = "none")
```

Some quick code with a simple Wilcoxon test, not incorporating the variable of experimental date.

```{r}
# boxplot with statistics added, basic wilcox test...

# list of comps for plots
my_comparisons = list(c("LD", "LL"))

# start plot
perf_data %>% 
  ggplot(aes(y = nym_per_aph, x = treat, fill = treat)) + 
  geom_violin() + 
  geom_jitter(width = 0.05) + 
  geom_boxplot(width=0.1, color="black", alpha=0.2, outlier.shape = NA) + 
  theme_classic() + 
  theme(legend.position = "none") + 
  xlab(NULL) + 
  ylab("Nymphs per Adult Aphid") +
  stat_compare_means(method = "wilcox.test", 
                     comparisons = my_comparisons, 
                     label = "p.signif",
                     method.args = list(exact = FALSE),
                     bracket.size = 0.75,       #  thickens  bracket line
                     size = 6,               # enlarges the asterisks
                     label.y = 17) +         # bumps bracket up 
  stat_compare_means(method = "wilcox.test", 
                     label.y.npc = "bottom", 
                     label.x.npc = "left",
                     size = 4,
                     method.args = list(exact = FALSE, conf.int = T, correct = T))
```

Updated modeling with a gamma glm. Including if the effect is true when also controlling for experimental date.
```{r}
# for gamma glm
library(MASS)
library(DHARMa)

# factorize
perf_data$Experiment = factor(perf_data$Experiment)

# see quick gamma glm. including both treat + experimental date.
full_mod = glm(nym_per_aph ~ treat + Experiment, data = perf_data, family = Gamma(link = "log"))

# summary
summary(full_mod)

# dharma
res_full_mod = simulateResiduals(fittedModel = full_mod, plot = TRUE)
```

Making an updated violin plot with significance data overlaid

```{r}
library(dplyr)
library(ggplot2)
# grab models with tidy()
library(broom)
# for pretty plots
library(ggprism) 

# factorize
perf_data = perf_data %>%
  mutate(
    treat = factor(treat, levels = c("LD", "LL")),
    Experiment = factor(Experiment)
  )

# fit the gamma
full_mod = glm(nym_per_aph ~ treat + Experiment, data = perf_data, family = Gamma(link = "log"))

# p-value extraction
perf_model_stats = broom::tidy(full_mod)

# filter for the treatment
treat_p_value = perf_model_stats %>%
  filter(grepl("treatLL", term)) %>%
  pull(p.value)

# sig levels
sig_stars = case_when(
  treat_p_value < 0.001 ~ "***",
  treat_p_value < 0.01  ~ "**",
  treat_p_value < 0.05  ~ "*",
  TRUE                  ~ "ns"
)

# the text gap and plot ceiling using 18 as the scaling anchor.
bracket_y = 18
text_y    = bracket_y + (18 * 0.02)  # maintains the same relative upward spacing for asterisks
limit_y   = bracket_y + (18 * 0.08)  # maintains the same proportional upper panel cushion

# start plot
p_perf = ggplot(perf_data, aes(x = treat, y = nym_per_aph, fill = treat)) +
  geom_violin(alpha = 1, color = NA, scale = "width", adjust = 0.8) + 
  geom_boxplot(width = 0.5, color = "grey10", outlier.shape = NA, alpha = 0.5) +
  geom_point(aes(shape = treat), 
             position = position_jitter(width = 0.08),
             color = "black", alpha = 0.5, size = 2) +
  # set shapes
  scale_shape_manual(values = c("LD" = 21, "LL" = 22)) +
  # apply tol bright palette
  scale_fill_manual(values = c("LD" = "#0077BB", "LL" = "#EE7733")) +
  # set labels
  labs(
    x = NULL, 
    y = "Nymphs per Adult Aphid"
  ) +
  theme_prism() +
  theme(legend.position = "none")

# add model p-values
if (treat_p_value < 0.05) {
  p_perf = p_perf + 
    # horizontal line at exactly 18
    annotate("segment", x = 1, xend = 2, y = bracket_y, yend = bracket_y, color = "black", linewidth = 0.5) +
    # add asterisk
    annotate("text", x = 1.5, y = text_y, label = sig_stars, size = 6, fontface = "bold") +
    # update panel borders
    coord_cartesian(ylim = c(0, limit_y))
}

# show fig
print(p_perf)
```

Save the plot to disc as svg.

```{r}
# save to svg
    # ggsave(
    #   "./figures_performance/perf_assay_stats.svg",
    #   plot = p_perf,
    #   width = 4, 
    #   height = 4,
    #   unit = "in"
    #        )
```

saving the coefficient tables to word doc for supp tables

```{r}
library(dplyr)
library(broom)
library(car)
library(flextable) 
library(officer)

# define the stripped-down formatting function
format_model_to_flextable_perf = function(model) {
  
  # extract coeff
  tidy_coef = broom::tidy(model)
  
  # extract anova
  anova_test = tryCatch({
    car::Anova(model, type = "III") %>% 
      broom::tidy() %>%
      filter(term != "(Intercept)") #keeps it clean
  }, error = function(e) {
    NULL
  })
  
  # start flextable

  make_basic_ft = function(df) {
    if(is.null(df)) return(NULL)
    
    flextable(df) %>%
      colformat_double(digits = 2) %>% 
      theme_booktabs() %>% 
      autofit()
  }
  
  # apply the helper function
  coef_ft  = make_basic_ft(tidy_coef)
  anova_ft = make_basic_ft(anova_test)
  
  return(list(coefs = coef_ft, anova = anova_ft))
}

# set dir and start saving
dest_dir = file.path(getwd(), "tables_performance")
if(!dir.exists(dest_dir)) dir.create(dest_dir)

# define the variable name for  titles
var_name = "Nymphs per Adult Aphid"

# format the tables using mode
tables = format_model_to_flextable_perf(full_mod)
file_path = file.path(dest_dir, "Performance_Assay_Results.docx")

# start blank Word Document
doc = read_docx()

#  attach text titles using body_add_par, then the tables
if(!is.null(tables$anova)) {
  doc = body_add_par(doc, paste("Type III ANOVA:", var_name), style = "Normal")
  doc = body_add_flextable(doc, value = tables$anova)
  doc = body_add_par(doc, "") 
}

if(!is.null(tables$coefs)) {
  doc = body_add_par(doc, paste("Model Coefficients:", var_name), style = "Normal")
  doc = body_add_flextable(doc, value = tables$coefs)
  doc = body_add_par(doc, "") 
}

# save the document
print(doc, target = file_path)
# debug print
print("Successfully saved Performance Assay Word document!")
```





Adult/ total aphid below likely wont need
##############################

total adults + nymphs

```{r}
library(dplyr)
library(ggplot2)
# grab models with tidy()
library(broom)
# for pretty plots
library(ggprism) 

# factorize
perf_data = perf_data %>%
  mutate(
    treat = factor(treat, levels = c("LD", "LL")),
    Experiment = factor(Experiment)
  )

# fit the Poisson GLM
full_mod_poisson = glm(total_aphids ~ treat + Experiment, family = poisson, data = perf_data)

# summary
summary(full_mod_poisson)

# dharma diagnostics to confirm overdispersion is handled
res_full_mod_pois = simulateResiduals(fittedModel = full_mod_poisson, plot = TRUE)

# p-value extraction (broom handles standard glms perfectly too)
perf_model_stats_poisson = broom::tidy(full_mod_poisson)

# filter for the treatment safely by searching for "treat" broadly
treat_p_value = perf_model_stats_poisson %>%
  filter(grepl("treat", term)) %>%
  pull(p.value)

# If multiple terms matched (e.g., if you had treatLL and treatSomethingElse), 
# grab the first treatment effect row
treat_p_value = treat_p_value[1] 

# Handle the case where the term vector might still be empty/NA safely
if (length(treat_p_value) == 0) treat_p_value = NA

# sig levels
sig_stars = case_when(
  is.na(treat_p_value)  ~ "ns",
  treat_p_value < 0.001 ~ "***",
  treat_p_value < 0.01  ~ "**",
  treat_p_value < 0.05  ~ "*",
  TRUE                  ~ "ns"
)

# dynamically set the text gap and plot ceiling using the max value 
max_val = max(perf_data$total_aphids, na.rm = TRUE)
bracket_y = max_val * 1.05
text_y    = bracket_y + (max_val * 0.02)  
limit_y   = bracket_y + (max_val * 0.08)  

# start plot
p_total = ggplot(perf_data, aes(x = treat, y = total_aphids, fill = treat)) +
  geom_violin(alpha = 1, color = NA, scale = "width", adjust = 0.8) + 
  geom_boxplot(width = 0.5, color = "grey10", outlier.shape = NA, alpha = 0.5) +
  geom_point(aes(shape = treat), 
             position = position_jitter(width = 0.08),
             color = "black", alpha = 0.5, size = 2) +
  scale_shape_manual(values = c("LD" = 21, "LL" = 22)) +
  scale_fill_manual(values = c("LD" = "#0077BB", "LL" = "#EE7733")) +
  labs(
    x = NULL, 
    y = "Total Aphids"
  ) +
  theme_prism() +
  theme(legend.position = "none")

# add model p-values safely
if (!is.na(treat_p_value)) { 
  p_total = p_total + 
    annotate("segment", x = 1, xend = 2, y = bracket_y, yend = bracket_y, color = "black", linewidth = 0.5) +
    annotate("text", x = 1.5, y = text_y, label = sig_stars, size = 6, fontface = "bold") +
    coord_cartesian(ylim = c(0, limit_y))
} else {
  # fallback if the model completely dropped the term
  p_total = p_total + 
    annotate("segment", x = 1, xend = 2, y = bracket_y, yend = bracket_y, color = "black", linewidth = 0.5) +
    annotate("text", x = 1.5, y = text_y, label = "ns", size = 6, fontface = "bold") +
    coord_cartesian(ylim = c(0, limit_y))
}

print(p_total)

# save to svg
    # ggsave(
    #   "./figures_performance/perf_assay_stats_TotalAphids.svg",
    #   plot = p_total,
    #   width = 4, 
    #   height = 4,
    #   unit = "in"
    #      )
```


total nymphs

```{r}
library(dplyr)
library(ggplot2)
library(broom)
library(ggprism) 

# factorize
perf_data = perf_data %>%
  mutate(
    treat = factor(treat, levels = c("LD", "LL")),
    Experiment = factor(Experiment)
  )

# fit the Poisson GLM for nymphs
mod_nymph_poisson = glm(nymph ~ treat + Experiment, family = poisson, data = perf_data)

# summary
summary(mod_nymph_poisson)

# dharma diagnostics
# res_nymph_pois = simulateResiduals(fittedModel = mod_nymph_poisson, plot = TRUE)

# p-value extraction
nymph_model_stats = broom::tidy(mod_nymph_poisson)

# filter for the treatment safely by searching for "treat" broadly
treat_p_value = nymph_model_stats %>%
  filter(grepl("treat", term)) %>%
  pull(p.value)

# grab the first treatment effect row if multiple matched
treat_p_value = treat_p_value[1] 

# handle the case where the term vector might still be empty/NA safely
if (length(treat_p_value) == 0) treat_p_value = NA

# sig levels
sig_stars = case_when(
  is.na(treat_p_value)  ~ "ns",
  treat_p_value < 0.001 ~ "***",
  treat_p_value < 0.01  ~ "**",
  treat_p_value < 0.05  ~ "*",
  TRUE                  ~ "ns"
)

# dynamically set the text gap and plot ceiling using the max value of nymphs
max_val = max(perf_data$nymph, na.rm = TRUE)
bracket_y = max_val * 1.05
text_y    = bracket_y + (max_val * 0.02)  
limit_y   = bracket_y + (max_val * 0.08)  

# start plot
p_nymph = ggplot(perf_data, aes(x = treat, y = nymph, fill = treat)) +
  geom_violin(alpha = 1, color = NA, scale = "width", adjust = 0.8) + 
  geom_boxplot(width = 0.5, color = "grey10", outlier.shape = NA, alpha = 0.5) +
  geom_point(aes(shape = treat), 
             position = position_jitter(width = 0.08),
             color = "black", alpha = 0.5, size = 2) +
  scale_shape_manual(values = c("LD" = 21, "LL" = 22)) +
  scale_fill_manual(values = c("LD" = "#0077BB", "LL" = "#EE7733")) +
  labs(
    x = NULL, 
    y = "Total Nymphs"
  ) +
  theme_prism() +
  theme(legend.position = "none")

# add model p-values safely
if (!is.na(treat_p_value)) { 
  p_nymph = p_nymph + 
    annotate("segment", x = 1, xend = 2, y = bracket_y, yend = bracket_y, color = "black", linewidth = 0.5) +
    annotate("text", x = 1.5, y = text_y, label = sig_stars, size = 6, fontface = "bold") +
    coord_cartesian(ylim = c(0, limit_y))
} else {
  # fallback if the model completely dropped the term
  p_nymph = p_nymph + 
    annotate("segment", x = 1, xend = 2, y = bracket_y, yend = bracket_y, color = "black", linewidth = 0.5) +
    annotate("text", x = 1.5, y = text_y, label = "ns", size = 6, fontface = "bold") +
    coord_cartesian(ylim = c(0, limit_y))
}

print(p_nymph)

# save to svg
# ggsave(
#   "./figures_performance/perf_assay_stats_TotalNymphs.svg",
#   plot = p_nymph,
#   width = 4, 
#   height = 4,
#   unit = "in"
# )
```

New plot combining the Total nymph + fecundity plots w/patchwork 

```{r}
library(patchwork)
# combined plot
combined_performance_plot = p_nymph + p_perf + 
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(size = 16, face = "bold"))

# take a peek before saving
print(combined_performance_plot)

# save to disc
ggsave(
  "./figures_performance/perf_assay_combined_nymphs.svg",
  plot = combined_performance_plot,
  width = 6,
  height = 4,
  unit = "in"
)
```


Quick word tables for the total nymph model

```{r}
# saving the total nymphs coefficient tables to word doc for supp tables
library(dplyr)
library(broom)
library(car)
library(flextable) 
library(officer)

# set dir (will use existing directory if already created)
dest_dir = file.path(getwd(), "tables_performance")
if(!dir.exists(dest_dir)) dir.create(dest_dir)

# define the variable name for titles
var_name_nymph = "Total Nymphs"

# format the tables using the Poisson model fit above
tables_nymph = format_model_to_flextable_perf(mod_nymph_poisson)
file_path_nymph = file.path(dest_dir, "Total_Nymphs_Assay_Results.docx")

# start blank Word Document
doc_nymph = read_docx()

#  attach text titles using body_add_par, then the tables
if(!is.null(tables_nymph$anova)) {
  doc_nymph = body_add_par(doc_nymph, paste("Type III ANOVA:", var_name_nymph), style = "Normal")
  doc_nymph = body_add_flextable(doc_nymph, value = tables_nymph$anova)
  doc_nymph = body_add_par(doc_nymph, "") 
}

if(!is.null(tables_nymph$coefs)) {
  doc_nymph = body_add_par(doc_nymph, paste("Model Coefficients:", var_name_nymph), style = "Normal")
  doc_nymph = body_add_flextable(doc_nymph, value = tables_nymph$coefs)
  doc_nymph = body_add_par(doc_nymph, "") 
}

# save the document
print(doc_nymph, target = file_path_nymph)
# debug print
print("Successfully saved Total Nymphs Assay Word document!")
```

